---
title: "Get putative enhancers (Enriched Annotations - Peaks vs. Whole Genome)"
author: "Miriam"
date: "7/14/2020, updated 17th August 2020, 24th August,7th Sept,6th Nov,10 Dec, 2 november  2022"
output: html_document
---

this schript is updated to new data
This script should do:  
(1) define new annotation definitions, then make a plot to look where the whole genome against the peaks re enriched (in which of the new annotations)  
(2) seperate the peaks into putative promoters, putative enhancers and CDS
(3) store that data  


###  (1) Find out which genetic regions are enriched in the DHS peaks compared to the whole genome   
```{r,echo=F,message=F, warning=F}
#read in the packages
library(GenomicRanges)
library(rtracklayer)
library(ChIPpeakAnno)
library(genomation)
library(BSgenome)
library(seqinr)
library(plyranges)
library(Biostrings)
library(reshape)
library(ggplot2)
library(knitr)
library(GenomicScores)
source("functions_regulation.R")
```


```{r,echo=F,message=F, warning=F}

#read in the files 
peaks <- read.table("data_from_novosparc/16dec2022_allgenes_doub_yolk_removed/peaks.txt",stringsAsFactors = F)
peaks <- makeGRangesFromDataFrame(reshape2::colsplit(peaks$V1,"-",names = c("chr","start","stop"))) 

#peaks <- readBed("data_for_novosparc/atac/atac_ranges.bed")
gtf <- import.gff("input/Drosophila_melanogaster.BDGP6.28.100.chr.gtf") #
fa <-"input/Drosophila_melanogaster.BDGP6.28.dna_rm.toplevel_nowhite.fa"

#which genes are expressed?
genes_exp <- read.table("data_from_novosparc/16dec2022_allgenes_doub_yolk_removed/genes.txt",stringsAsFactors = F)$V1

dir_out <-"output/annotate_ATAC_peaks/" #folder where all the output will be stored
if (!dir.exists(dir_out)) {dir.create(dir_out)}
```


```{r,echo=F,message=F, warning=F}
#subset to main chromosoms and feed the length information of the fasta file into the gtf file
gtf <- keepStandardChromosomes(gtf,pruning.mode = "coarse")
fasta<- seqinr::read.fasta( file = fa,seqtype = "DNA")
fasta<- fasta[1:7]
seqinfo(gtf)@seqlengths <-getLength(fasta) # add the chromosom length to the gtf file

seqlevelsStyle(peaks)<- "NCBI"
```


Since we dont want to compute bp's twice, we define our regions as the following and only stand-UNspecific: 

- "CDSplus" = [union of gtf:CDS and gtf:stop-codons and gtf:tRNA and gtf:snRNA and gtf:snoRNA]  
- "5'-UTR" = union of ([gtf:genes - "CDS"] overlap [gtf:5'-UTR])  
- "3'-UTR = union of ([gtf:genes - "CDS"] overlap [gtf: 3'-UTR])  
- "introns" = [gtf:genes] - union of ("CDS" + "5'-UTR" + "3'-UTR") (also pre-miRNA, ncRNA, lncRNA andpseudogenes are included here)  
- "intergenic" = [genom - masked] - [gtf:genes]
- "promoters" = first transcribed bp =/- 100 bp
- "intron_1" = the intron closest to the fiveprime end  

```{r,echo=F,warning=F}
#compute the "new" definitions for the gtf file annotations
masked_bps <-vmatchPattern("N",readDNAStringSet(fa, format = "fasta")[1:7])#get masked regions
MASKED <- GRanges(seqnames=NULL,ranges=NULL,strand=NULL)
for (i in names(masked_bps)){ 
  MASKED = append(MASKED, GRanges(seqnames = i,ranges = reduce(masked_bps[[i]]),strand = "*"))
} 

GENOM_wMA <- GRanges(seqnames=names(fasta),ranges=IRanges(start = rep(1,7), end = getLength(fasta) ))
GENOM <- setdiff(GENOM_wMA, MASKED) #the finally used genom witout the repetivie regions (cut out the masked)
GENES <- setdiff(reduce_ranges(gtf[gtf$type=="gene"]),MASKED) # there are also masked regions in the introns
INTERGENIC <- setdiff(GENOM, GENES)
CDS <-setdiff(reduce_ranges(gtf[gtf$type=="stop_codon" | gtf$type=="CDS" | gtf$type=="snoRNA" | gtf$type=="tRNA" | gtf$type=="snRNA"]),MASKED)
FIVEPRIME <-pintersect(findOverlapPairs(reduce_ranges(gtf[gtf$type=="five_prime_utr"]),GenomicRanges::setdiff(GENES,CDS)))
THREEPRIME <-pintersect(findOverlapPairs(reduce_ranges(gtf[gtf$type=="three_prime_utr"]),setdiff(GENES,append(CDS,FIVEPRIME))))
INTRON <- reduce_ranges(GenomicRanges::setdiff(GENES,reduce(append(append(CDS,FIVEPRIME),THREEPRIME))))
  
#additionally our definition of promoters (which is the first base of TSS =/- 100 bp)
fiveprimes <- gtf[gtf$type=="five_prime_utr"]
id <- fiveprimes$transcript_id; index<- logical(length(id)); last<-""
for (i in 1:length(id)){
  if (id[i] == last){last<-id[i]} else if (id[i] != last){index[i]<-TRUE; last<-id[i]}
} 
fiveprimes <- fiveprimes[index]
PROMOTERS <- reduce_ranges(flank(fiveprimes, 100,start= T,both=T))

#extra : look intro whether the first introns are enriched ( do this by taking last base of (first) 5'-UTR and then see if within 3 bp (start))
#the introns which split the 5'-UTRs
a <- fiveprimes
b <-resize(a,1,fix="end")
intron_a<-subsetByOverlaps(INTRON, b,maxgap = 0)
#and the ones which are behind the first CDS
a <- fiveprimes
b <-resize(a,1,fix="end")
c <- subsetByOverlaps(CDS, b,maxgap = 0)
intron_b <- subsetByOverlaps(INTRON,c ,maxgap = 0)
INTRON_1<- append(intron_a,intron_b) # all introns which are closest to the 5'-primeend


#print progress:
an_groups <- c("GENOM","GENES","CDS","FIVEPRIME","THREEPRIME","INTRON","INTERGENIC","PROMOTERS","INTRON_1")
for (i in (1:length(an_groups))){
  print(paste("The annotation group:",an_groups[i], "has",length(get(an_groups[i])),"objects"))
}

#write_bed(CDS, file=paste0(dir_out,"CDS.bed"))
#write_bed(INTERGENIC, file=paste0(dir_out,"intergenic.bed"))
#write_bed(PROMOTERS, file=paste0(dir_out,"promoters.bed"))
#write_bed(FIVEPRIME, file=paste0(dir_out,"fiveprime.bed"))
#write_bed(THREEPRIME, file=paste0(dir_out,"threeprime.bed"))

```

Compute the number of basepairs for each annotation group 
```{r,echo=F}
#make a summarytable
TAB <-data.frame(matrix(nrow= 4, ncol=8))
row.names(TAB) = c("genome_dm6","genome_percent","peaks","peaks_percent")
colnames(TAB) = c("bp","CDS","intron","intergenic","fiveprime","threeprime","promoters","intron_1")

#compute the sum of bp for the whole genome
TAB["genome_dm6","bp"] <- sum(width(GENOM))
TAB["genome_dm6","CDS"] <- sum(width(CDS))
TAB["genome_dm6","intron"] <- sum(width(INTRON))
TAB["genome_dm6","intergenic"] <-sum(width(INTERGENIC))
TAB["genome_dm6","fiveprime"] <- sum(width(FIVEPRIME))
TAB["genome_dm6","threeprime"] <-  sum(width(THREEPRIME))
TAB["genome_dm6","promoters"] <-  sum(width(PROMOTERS))
TAB["genome_dm6","intron_1"] <-  sum(width(INTRON_1))

#now calculate where the peaks fall in but only the overlapping bp 
TAB["peaks","bp"]<-sum(width(peaks))

#bp -peaks in CDS (only the overlap!)
TAB["peaks","CDS"] <- sum(width(pintersect(findOverlapPairs(peaks,CDS))))
TAB["peaks","intergenic"] <- sum(sum(width(peaks))-sum(width(pintersect(findOverlapPairs(peaks,GENES)))))
TAB["peaks","fiveprime"] <- sum(width(pintersect(findOverlapPairs(peaks,FIVEPRIME))))
TAB["peaks","threeprime"] <- sum(width(pintersect(findOverlapPairs(peaks,THREEPRIME))))
TAB["peaks","intron"] <- sum(width(pintersect(findOverlapPairs(peaks,INTRON))))
TAB["peaks","promoters"] <- sum(width(pintersect(findOverlapPairs(peaks,PROMOTERS))))
TAB["peaks","intron_1"] <- sum(width(pintersect(findOverlapPairs(peaks,INTRON_1))))

#compute the percentages
for (i in (1:dim(TAB)[2])){
  TAB["genome_percent",i] <- (TAB["genome_dm6",i]/TAB["genome_dm6","bp"])*100
}#for whole genome

for (i in (1:dim(TAB)[2])){
  TAB["peaks_percent",i] <- (TAB["peaks",i]/TAB["peaks","bp"])*100
}# for peaks

#check if correct
if (sum(TAB[1,2:6])== TAB[1,1]){
  print("TRUE: intron + CDS + intergenic + fiveprime + threeprime summs up to 100!")
} else {print("something does not match up");print(sum(TAB[1,2:6])); print(TAB[1,1])}

if (sum(TAB[3,2:6])== TAB[3,1]){
  print("TRUE: intron + CDS + intergenic + fiveprime + threeprime summs up to 100!")
  print(paste(round(TAB[3,1]/TAB[1,1]*100,2),"% of the unmasked genome is covered by the peaks we found"))
  } else {print("something does not match up");print(sum(TAB[3,2:6])); print(TAB[3,1])}

#print table
kable(round(TAB,1),caption = "Numbers of bp for genome vs ATAC peaks")
```


## Figure:

- CDS + intron + intergenic + fiveprime + threeprime = whole genome  
- whole genome = 137 Mio. bp - 29 Mio. masked regions  
- the promoters are our own defintion as : first base of fiveprime =/- 100bp  
- intron_1 stands for the subset of introns, which are the closest to the first base of fiveprime  

```{r,echo=F}
barplot(as.matrix(TAB[c(2,4),c(2:8)]),beside=T,main = "Basepair count peaks vs whole genome", las=2,col = c("green","blue"))
legend("topright",c("Genome", "ATAC peaks"),fill = c("green","blue"))


```



## (2) seperate the peaks into putative promoters, putative enhancer

divide the peaks into  

- putative promoters: min. 100 bp overlap with PROMOTERS (first transcript bp +/- 100)  
- CDS: min 100bp overlap with CDS
- putative enhancers: no overlap with PROMOTERS or CDS


```{r,echo=F}
#get a granges with indexes of 
peaks_t <- peaks

peaks_t$promoters <- FALSE
peaks_t$promoters[queryHits(findOverlaps(peaks_t,PROMOTERS,minoverlap = 100L))] <- TRUE
peaks_t$CDS <- FALSE
peaks_t$CDS[queryHits(findOverlaps(peaks_t,CDS,minoverlap = 100L))] <- TRUE
peaks_t$enhancers <- FALSE
#peaks$enhancers[queryHits(findOverlaps(peaks,not_CDS, minoverlap = 100L))]  <- TRUE
peaks_t$enhancers[! (peaks_t$promoters | peaks_t$CDS )] <- TRUE #get all which ar not yet defined
seqlevelsStyle(peaks_t)<-"UCSC"

#give the name
peaks_t$peak_ID <- paste0("peak_",c(1:length(peaks_t)))
```

### lets add some meta info to the peaks
```{r}

### give them a fake pseudo bulk score from the bigwigs
peaks_t$pseudo_score <- max_per_bigwig_peak(bigwigfile1 = "input/multiome/atac_cut_sites_sampleA.bigwig",bigwigfile2 = "input/multiome/atac_cut_sites_sampleB.bigwig",granges = peaks_t)

## overlaps with Vienna Tiles from the Stark lab 
VT_CAD4 <- import.bed("input/CAD4_plus_vienna_dm6.bed") #load the Vienna Tile
VT <- VT_CAD4[grep("^VT",VT_CAD4$name),]
peaks_t$overVT <- FALSE #generate a column
peaks_t$overVT[overlapsAny(peaks_t,VT)] <- TRUE
hits <- findOverlaps(peaks_t,VT) #overlap over 90% and which VT
overlaps <- pintersect(peaks_t[queryHits(hits)], VT[subjectHits(hits)])
percentOverlap <- width(overlaps) / width(peaks_t[queryHits(hits)])
hits_90 <- hits[percentOverlap >0.9] 
peaks_t$overVT_90 <- "NO" #add it to the table
peaks_t$overVT_90[queryHits(hits_90)] <- VT$name[subjectHits(hits_90)] #add the VT name


#add the nearest Genes 
seqlevelsStyle(gtf) <- "UCSC"
five_prime <- gtf[gtf$type=="five_prime_utr",] #subset to startcodons 
five_prime_availGenes <- five_prime[five_prime$gene_id %in% genes_exp,]

#add the nearest Gene
peaks_t$closGene1 <- five_prime$gene_id[GenomicRanges::nearest(peaks_t,five_prime)] #add the 1
peaks_t$closGene1_name <- five_prime$gene_name[GenomicRanges::nearest(peaks_t,five_prime)] #add the name

peaks_t$closGene2 <- c("id") #make an empty column for 2
peaks_t$closGene2_name <- c("name")
peaks_t$closGene3 <- c("id") #make an empty column for 3
peaks_t$closGene3_name <- c("name")


for (i in (1:length(peaks_t))){
  #meta_enhancers[i]
  #closest <- meta_enhancers$closGene1[i]
  #Closest Gene 1
  ind <- c()
  closGene1 <- peaks_t$closGene1[i]
  
  #closest gene 2
  sub_five_prime <- five_prime[!five_prime$gene_id==closGene1] #discard closest
  peaks_t$closGene2[i] <- sub_five_prime$gene_id[GenomicRanges::nearest(peaks_t[i],sub_five_prime)] #add the 2 closest gene-id
  peaks_t$closGene2_name[i] <- sub_five_prime$gene_name[GenomicRanges::nearest(peaks_t[i],sub_five_prime)] #add the gene_name
  closGene2 <-peaks_t$closGene2[i]
  #closest gene 3
  sub_sub_five_prime <- sub_five_prime[!sub_five_prime$gene_id==closGene2] #discard 1 and 2 closest
  peaks_t$closGene3[i] <- sub_sub_five_prime$gene_id[GenomicRanges::nearest(peaks_t[i],sub_sub_five_prime)] #add the 3 closest
  peaks_t$closGene3_name[i] <- sub_sub_five_prime$gene_name[GenomicRanges::nearest(peaks_t[i],sub_sub_five_prime)] #add the gene_nam
} 


```

```{r,echo=F}
#make into dataframe and give IDs!
df_peaks_t <- as.data.frame(peaks_t)
df_peaks_t$name <- paste(df_peaks_t$seqnames,df_peaks_t$start,df_peaks_t$end, sep = '-')
```

```{r,echo=F}
#split up in dataframes
df_peaks_promoters <- df_peaks_t[df_peaks_t$promoters,]
df_peaks_promoters$ID <- paste0("prom_",c(1:nrow(df_peaks_promoters)))

df_peaks_CDS <- df_peaks_t[ !df_peaks_t$promoters & df_peaks_t$CDS,]
df_peaks_CDS$ID <- paste0("CDS_",c(1:nrow(df_peaks_CDS)))

df_peaks_enhancers <- df_peaks_t[df_peaks_t$enhancers,]
df_peaks_enhancers$ID <- paste0("enh_",c(1:nrow(df_peaks_enhancers)))

#add info of IDS to big df
df_peaks_t$ID<- "ID"
df_peaks_t$annot<- "anno"
df_peaks_t$ID[df_peaks_t$peak_ID %in% df_peaks_promoters$peak_ID] <- df_peaks_promoters$ID
df_peaks_t$annot[df_peaks_t$peak_ID %in% df_peaks_promoters$peak_ID] <- "promoter"
df_peaks_t$ID[df_peaks_t$peak_ID %in% df_peaks_CDS$peak_ID] <- df_peaks_CDS$ID
df_peaks_t$annot[df_peaks_t$peak_ID %in% df_peaks_CDS$peak_ID] <- "CDS"
df_peaks_t$ID[df_peaks_t$peak_ID %in% df_peaks_enhancers$peak_ID] <- df_peaks_enhancers$ID
df_peaks_t$annot[df_peaks_t$peak_ID %in% df_peaks_enhancers$peak_ID] <- "enhancer"
```


```{r,echo=F}
#sve data as table and bedfile 
write.table(df_peaks_t,file = file.path(dir_out,"peaks_all.txt"),quote = F,row.names = F)
write_bed(df_peaks_t,file.path(dir_out,"peaks_all.bed"))

write.table(df_peaks_CDS,file = file.path(dir_out,"peaks_CDS.txt"),quote = F,row.names = F)
write_bed(df_peaks_CDS,file.path(dir_out,"peaks_CDS.bed"))
write.table(df_peaks_enhancers,file = file.path(dir_out,"peaks_enhancers.txt"),quote = F,row.names = F)
write_bed(df_peaks_enhancers,file.path(dir_out,"peaks_enhancers.bed"))
write.table(df_peaks_promoters,file = file.path(dir_out,"peaks_promoters.txt"),quote = F,row.names = F)
write_bed(df_peaks_promoters,file.path(dir_out,"peaks_promoters.bed"))
```











